---
title: "Problem Set 2"
author: "Anastasia Bernat"
date: "4/1/2020"
output: html_document
---
# Problem Set 2 | 1 April 2020 - 8 April 202 at 10 AM

**Problem 11. From PSET 1**

We didn't get to this problem but go ahead and try it this time! Write an R program to convert temperatures to and from celsius, fahrenheit. You may need to look back to Lesson 3 to revisit if statements.

```{r}

```

## While Loops

From lesson 5, we learned that while loops take this form:

while (boolean-expression) {
  block of instructions
}

Also, if you don't know already, if you press cmd + option + i, you can create new R blocks.

**Problem 1**

When given variables K = 5 and I = -2, write a while loop that prints the sum of I and K after repeatedly adding 2 to I and subtracting 1 from K **until** I is greater than K. 

```{r}
K <- 5
I <- -2
```

**Problem 2**

When given the variable int = 4L, write a while loop that subtracts 1 from int and then calculates its cubed value **until** int becomes negative. Afterwards, **plot** the cubed values. 

```{r}
int = 4L
```

**Problem 3**

Print the following pattern using a while loop:

*
**
***
****

You will need to use the rep() function and the cat() function to be able to correctly print out this pattern.

```{r}

```


**Problem 4**

Print the multiplication table of 50 (up until it's 10th multiple of 50) using a while loop.

```{r}

```

**Problem 5**

Print a vector with the factors of 5 (till 50) using a while loop. The vector should print these values backwards as such: 

[1] 50 25 10  5

```{r}

```

## Repeat Loops 

From lesson 5, we learned that repeat loops take a slightly different form than while loops. The boolean expression is at the bottom instead of the beginning of the loop:

repeat {
  block of instructions
  (boolean-expression) {
    break
  }
}

Let's try some repeat loop problems

**Problem 6**

Using a repeat loop, print out the following:

How many licks does it take to get to the center of a Tootsie Pop?
1
2
CRUNCH. The world may never know.

```{r}

```

**Problem 7**

Repeat problem 3 but using a repeat loop. Problem 3 said, print this following pattern:

*
**
***
****

```{r}

```

**Problem 8**

Starting at 1, print every odd number from 1-19.

```{r}
i <- 1

```

**Problem 9**

Using a repeat loop, count how many days have past since the new year. Read up on the library lubridate and the function as_date to understand what is happening. Here is one source but check out others as well: (https://www.rdocumentation.org/packages/lubridate/versions/1.7.4/topics/as_date)


```{r}
library(lubridate)

date <- as_date("2020-01-01")

```

