---
title: "Problem Set 4"
author: "Anastasia Bernat"
output: html_document
---
# Problem Set 4 | 8 April 2020 -15 April 202 at 10 AM

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Below there are 2  nested for loop questions, but please be sure to answer the questions that proceed the '#' in the R code blocks. 

**Problem 1.**

Write a nested for loop that will print the following output:

1
2 2
3 3 3
4 4 4 4

Hint: use the cat function to print correctly. 

```{r}
# Write nested loop here:





# -----------------------------------
# How many times was the outler loop executed?
# How many times was the inner loop executed? 

# Which numbers does the inner loop loop over on the third execution of the outler loop and why?  
```

**Problem 2.**

Count the characters in all these movie titles and print the results one by one to your screen, in this format:

"The title [movie_title] is [X] characters long."

Hint: Look into the stringr library and its functions to help you complete the task. 

```{r}
library(stringr) 
my_movies = list(c("Dead Poets Society", "Life is Beautiful"),
                 c("Radium Girls", "Parasite"),
                 c("Inception", "Interstellar"))

# Write nested loop here:






# -----------------------------------
# How many times was the outler loop executed? 
# How many times was the inner loop executed? 
```
