---
title: "Nesting Functions and Nesting Scripts"
author: "Anastasia Bernat"
output: html_document
---

```{r setup, include=FALSE}
setwd("/Users/anastasiabernat/Desktop/Rlessons/")
```

# Lesson 11 | 27 May 2020

## Functions within functions, scripts within scripts

In the last lesson, I showed you how to build a user-based function in R. They're great because you can create them to perform any repeated calculation or formula you need. However, functions can get pretty long at times and even a function's block of statements can also become repetitive. That's where the benefit of nested functions and nested scripts come in. By nesting funcitons and scripts, you can create a more readable and organized code. 

To start, let's look at some quick examples of nested functions and how to source a script in R.

## Nested Functions.

There are a few ways to structure them. Here are a couple ways:

(i) Call a function within another function.

(ii) Write a function within another function.

```{r}
# EXAMPLE 1: Call a function within another function

sq.root <- function(x) {
  root <- x**(1/2)
  return(root)
}

sq.root(sq.root(25))

# EXAMPLE 2: Call a function within another function

plot_roots <- function(nums) {
  plot(1, type="n", xlab="numbers", ylab="roots", xlim=c(0, max(nums)), ylim=c(0, sq.root(max(nums))))
  for (n in nums) {
    r <- sq.root(n)
    lines(n, r, type="p")
  }
}
plot_roots(c(1,2,3,4,5))
```

```{r}
# EXAMPLE 3: Write a function within another function.

num1 <- function(x) { # outer function
  num2 <- function(y) { # inner function
    return(x*y)
  }
  return(num2)
}

result <- num1(10)
print(result)
print(result(5))
```

**What happened here?** 

(i) The inner function, num2, has been defined inside the outer funciton called num1. 

(ii) To call num2 we need to first call num1. To call num1 do num1(10) and store it in a variable called result. If we print result, we'll see that it'll print the inner function because what is returned in the outer function was num2. 

(iii) To then call num2, we use the num2 function that we stored in the result variable and input a y value, which was 5 in the above example. So what gets calculated is 10*5, which equals 50. 

This may be hard to visualize or picture at first. For example, you might be wondering, where did the x=10 go then if it doesn't show up when we print(result)? This gets into what is known as the call stack. Here's a helpful site that breaks down the call stack through diagrams if you want to understand what's happening under the hood: https://swcarpentry.github.io/r-novice-inflammation/14-supp-call-stack/

### How to Source a Script in R.

**Terms:**

*Working directory (wd): a file path on your computer that sets the default location of any files you read into R, or save out of R.

*Forward slash: The separator in the path names is the backslash, which in R is the escape character in characters strings. So, use forward slashes instead of the backslashes.

**Functions:** 

getwd() - when you run this funciton in the console with no arguments it will print the path of the working directory you are in

setwd() - when you input the path of the working direction you want to work in as a character, it will change the working directory to the folder you want to work in.

source() - when you input the filename or path of a file, URL, connection, or expressions directly you want to run in another script as a character, this function will accept and read your input directly.

**EXAMPLE 4 **

So, now let's pull up an old example - converting temperatures - and turn it into a function. Then, we will move it to another script and source the script in this file.

```{r}
############################################################################
# FUNCTION: convert_temp | converts any temperature into C, F, or K
# INPUT:  First, a temperature you want to convert as a string consisting of 
#         a value followed by its units. Then, the unit you want to convert
#         temp to as a string.
# OUTPUT: returns the converted temperature as a string consisting of a 
#         value and its units.
############################################################################

convert_temp <- function(temp, convert_to) {

  measure <- as.numeric(substr(temp,1, nchar(temp)-1))
  input_unit = substr(temp, nchar(temp), nchar(temp))
  
  if (input_unit == "C") {
    if (convert_to == "F") {
      converted_temp = (9 * measure) / 5 + 32
      output_unit = "Fahrenheit" }
    else if (convert_to == "K") {
      converted_temp = as.numeric(measure) + 273.15
      output_unit = "Kelvin"}
  }
  
  if (input_unit  == "F") {
    if (convert_to == "C") {
      converted_temp = (measure - 32) * 5 / 9
      output_unit = "Celsius" }
    else if (convert_to == "K") {
      converted_temp <- ((measure - 32) * 5/9) + 273.15
      output_unit = "Kelvin"}
    }
  
  if (input_unit  == "K") {
    if (convert_to == "C") {
      converted_temp = as.numeric(measure) - 273.15
      output_unit = "Celsius" }
    else if (convert_to == "F") {
      converted_temp <- ((measure - 273.15) * 9/5) + 32
      output_unit = "Kelvin"}
    }
  paste("The temperature in", output_unit, "is", converted_temp, "degrees.")
  
  converted_temp <- paste0(converted_temp, convert_to)
  return(converted_temp)
}
```

```{r}
source("converting-temps.R")                  # Sourcing a script 
convert_temp("25K", "C")                      # Calling the function
convert_temp(convert_temp("25K", "C"), "F")   # Calling a function within a function
```

