---
title: "Problem Set 3"
author: "Anastasia Bernat"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Set 3 | 8 April 2020 - 15 April 202 at 10 AM

**Problem 1. Review**

Let's play a game - snakes and ladders! 

Snakes and Ladders is an ancient Indian board game regarded today as a worldwide classic. It is played between two or more players on a board with numbered, gridded squares and populated by "ladders" and "snakes". These "ladders" and "snakes" connect two specific board squares together. If you land on a ladder after the role of a die, you jump up the board game and get closer to the final square. If you land on a snake, you fall down the board game.

How to play:

* Each player rolls the die in order to progress up the board game.

* If your counter lands at the bottom of a ladder, you can move up to the top of the ladder.

* If your counter lands on the head of a snake, you must slide down to the bottom of the snake.

* The first player to get to the last square on the board is the winner.


```{r}
# Snakes and Ladders 

# Step 1: Create Your Board with Snakes And Ladders




# Step 2: Initializations

# Step 3: Loop

# You will need your loop to allow you to 
## 1. move up or down for a snake or ladder
## 2. roll the dice
## 3. move by that rolled amount
## 4. once you get to the final square, game stops and print("Game over!")





```

**Problem 2. Nested Loops**

Use a nested-for loop to display the multiplication tables from 1-10.  

```{r}

```

